Abstractions and exemplars: the measure noun phrase alternation in German

### 2018-05-30 Latex/article.tex
Adapted MCMC script for introduction to habilitation.
Final touches after final acceptance.
Final final touches.

%\usepackage[utf8]{inputenc}
\RequirePackage[no-math]{fontspec}

@@

\usepackage[firstpage]{draftwatermark}
\SetWatermarkFontSize{24pt}
\SetWatermarkHorCenter{300pt}
\SetWatermarkColor[rgb]{0.9,0.2,0.2}
\SetWatermarkText{\textsf{Final revision. Quote as draft and to appear in COGL.}}

@@

% \runninglinenumbers
% \modulolinenumbers[1]
% \setlength\linenumbersep{20pt}

@@

  \author*[1]{Roland Schäfer}
  \runningauthor{Roland Schäfer}
  \affil[1]{Freie Universität Berlin}

@@

  \abstract{In this paper, an alternation in German measure noun phrases is examined under a varying-abstraction perspective.
    In a specific measure NP construction, the embedded kind-denoting noun either agrees in case with the measure noun (\textit{eine Tasse guter Kaffee} `a cup of good coffee') or it stands in the genitive (\textit{eine Tasse guten Kaffees}).
    Each of the two alternants is syntactically similar to a non-alternating construction.
    I propose a prototype model which assigns a common prototypical meaning to each of the alternants and its corresponding non-alternating construction.
    Based on this, I argue that lexical, morpho-syntactic, and stylistic features help to predict the choice of the alternant.
    A large corpus study is presented which supports this analysis.
    However, in addition to the prototype effects, an exemplar effect is also shown to influence the choice, namely the relative frequencies with which lemmas occur in the non-alternating constructions.
    I argue that allowing both prototype and exemplar effects is more adequate than following radical prototype or exemplar approaches.
    It is also verified in two experiments that the corpus-derived model corresponds to the behaviour of native speakers.
    The weak effect size of the experimental validation is discussed in the context of corpus-based cognitive linguistics and the validation of corpus-derived models.}
  \keywords{prototypes vs.\ exemplars, corpus methods and experimental validation, alternations, multilevel modeling, pseudo-partitives, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  \received{18 April 2017}
  \accepted{3 April 2018}
  \journalname{To appear in Cognitive Linguistics}
  %\journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

% For word counting.
% \raggedright
% \pagestyle{empty}

>>>>

\RequirePackage[no-math]{fontspec}

@@

@@

@@

  \author*[1]{Roland Schäfer}
  \runningauthor{Roland Schäfer}
  \affil[1]{Deutsche und niederländische Philologie, Freie Universität Berlin, email: roland.schaefer@fu-berlin.de}

@@

  \abstract{In this paper, an alternation in German measure noun phrases is examined under a varying-abstraction perspective.
    In a specific measure NP construction, the embedded kind-denoting noun either agrees in case with the measure noun (\textit{eine Tasse guter Kaffee} `a cup of good coffee') or it stands in the genitive (\textit{eine Tasse guten Kaffees}).
    Each of the two alternants is syntactically similar to a non-alternating construction.
    I propose a prototype model which assigns a common prototypical meaning to each of the alternants and its corresponding non-alternating construction.
    Based on this, I argue that lexical, morpho-syntactic, and stylistic features help to predict the choice of the alternant.
    A large corpus study is presented which supports this analysis.
    However, in addition to the prototype effects, an exemplar effect is also shown to influence the choice, namely the relative frequencies with which lemmas occur in the non-alternating constructions.
    I argue that allowing both prototype and exemplar effects is more adequate than following radical prototype or exemplar approaches.
    It is also verified in two experiments that the corpus-derived model corresponds to the behaviour of native speakers.
    The weak effect size of the experimental validation is discussed in the context of corpus-based cognitive linguistics and the validation of corpus-derived models.}
  \keywords{prototypes vs.\ exemplars, corpus methods and experimental validation, morpho-syntactic alternations, multilevel modeling, pseudo-partitives, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{}
  %\accepted{}
  %\journalname{}
  %\journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

### 2018-05-30 Latex/sections/1intro.tex
Final touches after final acceptance.
Final final touches.

\begin{center}
  Data and scripts: \url{https://github.com/rsling/measurenps}
\end{center}

@@

In cognitive linguistics, similarity-based categorisation is often seen as the central and conceptually sufficient approach to linguistic categorisation.
See \cite{Taylor2003} for a comprehensive treatment in terms of prototype theory and \cite{Taylor2012} for a detailed discussion of an exemplar-based approach.
\cite{DivjakArppe2013} is a rare example of a paper in cognitive linguistics where a synthesis between prototype and exemplar models is proposed.
Their corpus-based approach shows ``one way of systematically analyzing usage data as contained in corpora to yield a scheme, compatible with usage-based theories of language, by which the assumptions of both the prototype and exemplar theories can be operationalized'' \citep[267]{DivjakArppe2013}.
Their approach to implementing a varying abstraction model \citep[254--260]{DivjakArppe2013} is based on hierarchical clustering of annotated properties of sentences.
They cluster sentences containing Russian verbs of trying.
Then, they single out the one sentence from each cluster which scores the highest probability for any of the six \textit{try} verbs according to a polytomous regression model estimated on the same data.
The clusters are interpreted as intermediate-level exemplar-derived abstractions of typical contexts for these high-probability verbs (typically more than one cluster for each verb; \citealp[255--256]{DivjakArppe2013}).
The crucial difference between such data-driven corpus-based analyses and experiments in cognitive science (\citealp{DivjakArppe2013} use \citealp{VerbeemenEa2007} as their reference) is that cognitive research is based on experiments where subjects produce actual category assignments or similarity judgements, and in corpus studies, the categories and category membership are determined purely from existing data.
The experimental approach with reduced and\slash or artificial stimuli makes it much easier to examine very specific effects in the behaviour of the subjects.
While I am convinced that the results presented in \citet{DivjakArppe2013} are valid and important (especially given the previous and subsequent research the authors have conducted on the data, including experimental work presented in \citealp{DivjakEa2016}), any data set can be clustered to yield a certain number of clusters.
Thus, the corpus study alone does not ensure that the clusters emerging from the data correspond to any particular speaker's cognitive representation.

>>>>

@@

In cognitive linguistics, similarity-based categorisation is often seen as the central and conceptually sufficient approach to linguistic categorisation.
See \cite{Taylor2003} for a comprehensive treatment in terms of prototype theory and \cite{Taylor2012} for a detailed discussion of an exemplar-based approach.
\cite{DivjakArppe2013} is a rare example of a paper in cognitive linguistics where a synthesis between prototype and exemplar models is proposed.
Their corpus-based approach shows ``one way of systematically analyzing usage data as contained in corpora to yield a scheme, compatible with usage-based theories of language, by which the assumptions of both the prototype and exemplar theories can be operationalized'' \citep[267]{DivjakArppe2013}.
Their approach to implementing a varying abstraction model \citep[254--260]{DivjakArppe2013} is based on hierarchical clustering of annotated properties of sentences.
They cluster sentences containing Russian verbs of trying.
Then, they single out the one sentence from each cluster which scores the highest probability for any of the six \textit{try} verbs according to a polytomous regression model estimated on the same data.
The clusters are interpreted as intermediate-level exemplar-derived abstractions of typical contexts for these high-probability verbs (typically more than one cluster for each verb; \citealp[255--256]{DivjakArppe2013}).
The crucial difference between such data-driven corpus-based analyses and experiments in cognitive science (\citealp{DivjakArppe2013} use \citealp{VerbeemenEa2007} as their reference) is that cognitive research is based on experiments where subjects produce actual category assignments or similarity judgements, and in corpus studies, the categories and category membership are determined purely from existing data.
The experimental approach with reduced and\slash or artificial stimuli makes it much easier to examine very specific effects in the behaviour of the subjects.
While I am convinced that the results presented in \citet{DivjakArppe2013} are valid and important (especially given the previous and subsequent research the authors have conducted on the data, including experimental work presented in \citealp{DivjakGries2008,DivjakEa2016}), any data set can be clustered to yield a certain number of clusters.
Thus, the corpus study alone does not ensure that the clusters emerging from the data correspond to any particular speaker's cognitive representation.

### 2018-05-30 Latex/sections/3corpus.tex
Final final touches.

For this pre-study, each of the 2,365 combinations of measure noun lemma and kind noun lemma were queried in the \NACb\ and the \PGCd\ with a limit of 100 randomly chosen results for each single query.%
\footnote{The data package for this paper (including all concordances and R scripts) is made available under the DOI \doi{10.5281/zenodo.1254871}.}
From the resulting 35,766 sentences, 5,000 were randomly sampled for the statistical analysis.
All features described in Section~\ref{sec:variablesandannotation} were annotated except for the ones which do not apply in the non-alternating case (\textit{Kindattraction}, \textit{Measureattraction}, and \textit{Measurecase}).

>>>>

For this pre-study, each of the 2,365 combinations of measure noun lemma and kind noun lemma were queried in the \NACb\ and the \PGCd\ with a limit of 100 randomly chosen results for each single query.%
\footnote{The data package for this paper (including all concordances and R scripts) is made available under the DOI \url{https://dx.doi.org/10.5281/zenodo.1254871}.}
From the resulting 35,766 sentences, 5,000 were randomly sampled for the statistical analysis.
All features described in Section~\ref{sec:variablesandannotation} were annotated except for the ones which do not apply in the non-alternating case (\textit{Kindattraction}, \textit{Measureattraction}, and \textit{Measurecase}).

### 2018-05-30 Latex/sections/4experiments.tex
Final version.
Final final touches.

In retrospect, this stimulus was badly chosen because \textit{heiße Luft} `ado' (literally `hot air') is a fixed metaphorical expression.
This obviously influences the reactions of participants, and a revised and improved experiment might lead to a much better fit in future research.

@@

In principle, random slope for \textit{Modelprediction} varying by item could also remedy problems with individual stimuli at least partially.
Therefore, a model with random slopes for \textit{Modelprediction} varying by both random effects (\textit{Participant} and \textit{Item}) was specified and estimated.
The random slope for participants was added to comply with \citet[257]{BarrEa2013} who predict ``catastrophically high Type I error rates'' for experimental designs with within-subject manipulations if random effects structures are not kept maximal.
The coefficient of the fixed effect changed noticeably but not enough to change the interpretation (5.408 relative to an intercept of -1.304), and the marginal $R^2_m$ rises to 0.213 ($R^2_c=0.488$).
In line with expectations, the standard deviation in the random slopes for \textit{Item} is high at 5.996.
However, the covariance parameters were estimated at exactly -1, which is a clear sign that the variance-covariance matrix could not be estimated successfully.
The same was true for models with only an \textit{Item} and a \textit{Participant} random slope.
This is exactly the kind of model overparametrisation criticised in \citet{BatesEa2015a} and \citet{MatuschekEa2017}.
The available data are insufficient to estimate the parameters of the more complex model with varying slopes.%
\footnote{\citet[1]{BatesEa2015a} state: ``We show that failure to converge typically is not due to a suboptimal estimation algorithm, but is a consequence of attempting to fit a model that is too complex to be properly supported by the data, irrespective of whether estimation is based on maximum likelihood or on Bayesian hierarchical modelling with uninformative or weakly informative priors.
Importantly, even under convergence, over-parameterisation may lead to uninterpretable models.''
}

>>>>

In retrospect, this stimulus was badly chosen because \textit{heiße Luft} `ado' (literally `hot air') is a fixed metaphorical expression.
This obviously influences the reactions of participants, and a revised and improved experiment might lead to a much better fit in future research.%
\footnote{Since I am convinced that the statistical analysis of an experiment has to be determined as part of the experiment's design before the experiment is conducted, removing the data points corresponding to the problematic stimulus from the analysis is not an option.
Future experiments with improved stimuli is the proper route to take.}

@@

In principle, random slope for \textit{Modelprediction} varying by item could also remedy problems with individual stimuli at least partially.
Therefore, a model with random slopes for \textit{Modelprediction} varying by both random effects (\textit{Participant} and \textit{Item}) was specified and estimated.
The random slope for participants was added to comply with \citet[257]{BarrEa2013} who predict ``catastrophically high Type I error rates'' for experimental designs with within-subject manipulations if random effects structures are not kept maximal.
The coefficient of the fixed effect changed noticeably but not enough to change the interpretation (5.408 relative to an intercept of -1.304), and the marginal $R^2_m$ rises to 0.213 ($R^2_c=0.488$).
In line with expectations, the standard deviation in the random slopes for \textit{Item} is high at 5.996.
However, the covariance parameters were estimated at exactly -1, which is a clear sign that the variance-covariance matrix could not be estimated successfully.
The same was true for models with only an \textit{Item} or only a \textit{Participant} random slope.
This is exactly the kind of model overparametrisation criticised in \citet{BatesEa2015a} and \citet{MatuschekEa2017}.
The available data are insufficient to estimate the parameters of the more complex model with varying slopes.%
\footnote{\citet[1]{BatesEa2015a} state: ``We show that failure to converge typically is not due to a suboptimal estimation algorithm, but is a consequence of attempting to fit a model that is too complex to be properly supported by the data, irrespective of whether estimation is based on maximum likelihood or on Bayesian hierarchical modelling with uninformative or weakly informative priors.
Importantly, even under convergence, over-parameterisation may lead to uninterpretable models.''
}
